---
title: "Recommendation_spotify"
author: "Uday Gadge"
date: "2022-12-11"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Recommendation spotify

For this I am using my 2022 wrapped top songs as a base. Spotify API also provides artists that are similar to an aritst. I picked a few of my favourite artists and got artists similar to them. I can calculate distances of these songs from this library and the top songs of 2022.
path <- "C:/Users/gadge/OneDrive/Desktop/InfoStats_R/"
filename <- "Spotify_data.csv"
df <- read_csv(paste(path,filename,sep=""))
getting data

```{r,include=FALSE}
library(dplyr)
library(tidyverse)
library(gridExtra)
library(proxy)
library(ggforce)
```

```{r,include = FALSE}
path <- "C:/Users/gadge/OneDrive/Desktop/InfoStats_R/"
filename <- "Top_songs_2022.csv"
top_songs <- read_csv(paste(path,filename,sep=""))
```

```{r,include=FALSE}
filename <- "library_recommender.csv"
library <- read_csv(paste(path,filename,sep=""))
```

```{r}
colnames(top_songs)
```
```{r}
top_songs$album_release_year <- format(as.Date(top_songs$track.album.release_date,
                                        format="%Y-%m-%d"),"%Y")
top_songs <- top_songs %>% select(-track.artists,-track.explicit,-track.popularity,
                                  -track.album.release_date)
```

```{r}
head(top_songs)
```

```{r}
dim(top_songs)
dim(library)
```



```{r}
head(library)
```
```{r}
library <- library %>% select(track_id,artist_id,album_id,
                              track_name,artist_name, album_release_year, 
                              duration_ms, danceability, energy, loudness, 
                              key, mode, speechiness, acousticness, 
                              instrumentalness, liveness, valence, 
                              tempo)
head(library)
```
```{r}
top_songs <- top_songs %>% rename("track_id" = "track.id",
                     "track_name" = "track.name",
                     "duration_ms" = "track.duration_ms")
head(top_songs)
```

```{r}
song_features_num <- c("danceability","energy","loudness","acousticness","instrumentalness",
                   "liveness","valence","tempo","album_release_year")

song_features_cat <- c("key","mode")
```

```{r}
identifiers <- c("track_id","track_name")

top_songs <- top_songs %>% dplyr::select(any_of(c(identifiers,song_features_num,
                                                  song_features_cat)))
```

```{r}
library <- library %>% dplyr::select(any_of(c(identifiers,song_features_num,
                                                  song_features_cat)))
head(library)
```

```{r}
library <- library %>% filter(!(track_id %in% top_songs$track_id))
dim(library)
```

```{r}
top_songs$top_song <- "Yes"
library$top_song <- "No"

df <- rbind(top_songs,library)
```

```{r}
df <- data.frame(df)
head(df)
```


```{r}
df$album_release_year <- as.double(df$album_release_year)
head(df)
```
```{r}
sum(is.na(df))
```
```{r}
df <- na.omit(df)
```

```{r}
dim(df)
```


```{r}
for (col in song_features_num){
  #print(col)
  #df[,col] <- as.numeric(df[,col])
  df[,col] <- (df[,col] - min(df[,col]))/(max(df[,col])-min(df[,col]))
}
head(df)
```

```{r}
df$key <- as.factor(df$key)
df$mode <- as.factor(df$mode)

```


```{r}
modes <- df %>% 
  dplyr::group_by(mode) %>% 
  dplyr::summarize(mode_prob = (n())/(dim(df)[1]))
keys <- df %>% 
  dplyr::group_by(key) %>% 
  dplyr::summarize(key_prob = (n())/(dim(df)[1]))
modes
```

```{r}
df <- left_join(df,modes,by="mode")
df <- left_join(df,keys,by="key")
head(df)
```

```{r}
song_features_cat <- c("key_prob", "mode_prob")
a <- df[df$top_song == "Yes",c(song_features_num,song_features_cat)]
b <- df[df$top_song == "No",c(song_features_num,song_features_cat)]
class(a)
```
```{r}
head(a)
```
```{r}
distance_prob <- function(v1,v2){
  sqrt(sum((v1[1:8]-v2[1:8])^2) +
         sum((v1[9:11] == v2[9:11])*v1[9:11]*v2[9:11])+
         sum((v1[9:11] != v2[9:11])*(1-v1[9:11]*v2[9:11])))
}
```



```{r}
library(proxy)
mat <- proxy::dist(a,b,method = distance_prob)
```


```{r}
dim(mat)
```

```{r}
mat[1:3,2:5]
```

```{r}
which.min(mat[1,2:5])
```
```{r}
which.min(mat[1,2:5])-1
```


```{r}
similar <- c()
for (i in (1:dim(mat)[1])){
  similar <- c(similar, which.min(mat[i,]))
}
```

```{r}
similar
```
```{r}
top_song_matched <- df %>% filter(top_song == "Yes")
top_song_matched$similar_index <- similar
top_song_matched
```

```{r}
similar_song_id <- c()
similar_song_names <- c()
for (i in similar){
  similar_song_id <- c(similar_song_id,df$track_id[[i+98]])
  similar_song_names <- c(similar_song_names,df$track_name[[i+98]])
}
similar_song_names
```


```{r}
mean_distance <- c()
for (i in (1:dim(mat)[2])){
  mean_distance <- c(mean_distance, mean(mat[,i]))
}

```

```{r}
mean_distances <- data.frame(mean_distance, index = seq(1,length(mean_distance)))
head(mean_distances)
```

```{r}
head(arrange(mean_distances,mean_distances$mean_distance))
```
```{r}
top_30 <- (arrange(mean_distances,mean_distances$mean_distance))$index[1:30]
```

```{r}
top_30_song_id <- c()
top_30_song_names <- c()
for (i in top_30){
  top_30_song_id <- c(top_30_song_id,df$track_id[[i+98]])
  top_30_song_names <- c(top_30_song_names,df$track_name[[i+98]])
}
top_30_song_names

```
```{r}
pc <- prcomp(df[,song_features_num],center = TRUE)
```

```{r}
attributes(pc)
```

```{r}
print(pc)
```

```{r}
pca_df <- data.frame(predict(pc,df[,song_features_num]))
data_pca <- cbind(df,pca_df)
head(data_pca)
```

```{r}
data_pca %>% filter(top_song == "Yes" | track_id %in% similar_song_id) %>% 
  ggplot(aes(PC1,PC2,color = top_song)) + geom_point()+
  geom_text(aes(PC1,PC2,label = track_name),size = 1.5) + xlab("") +ylab("")+
  theme(axis.text.y=element_blank(),axis.text.x=element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("Individually Similar Songs")
```

```{r}
data_pca %>% filter(top_song == "Yes" | track_id %in% top_30_song_id) %>% 
  ggplot(aes(PC1,PC2,color = top_song)) + geom_point()+
  geom_text(aes(PC1,PC2,label = track_name),size = 1.5) + xlab("") +ylab("")+
  theme(axis.text.y=element_blank(),axis.text.x=element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("Overall Similar Songs")

```


