---
title: "Recommender_data"
author: "Uday Gadge"
date: "2022-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(dplyr)
library(tidyverse)
```

## Gathering data for recommender system

```{r}
library(spotifyr)
Sys.setenv(SPOTIFY_CLIENT_ID = '39fab145f77749eb9e00395cb262d536')
Sys.setenv(SPOTIFY_CLIENT_SECRET = '6e8c8e372ede44eb9f233a9485f50538')
Sys.setenv(redirect_uri = 'http://localhost:8888/callback')

access_token <- get_spotify_access_token()
```

I have choosen the 2022 wrapped top songs as a base for the recommender system. It has 100 tracks.

```{r}
playlist <- get_playlist("1HW8XyRWunaQJxLtJJIthD?si=hPBRxMZ-RvqQhM_yznxsJQ&nd=1")$tracks
```

```{r}
top_songs_2022 <- playlist$items
```

Get only required columns

```{r}
top_songs_2022 <- top_songs_2022 %>% select(track.id,track.name,track.artists,
                                            track.duration_ms,track.explicit,
                                            track.popularity,track.album.release_date)
```

```{r}
top_songs_2022
```

Get the audio features of the song

```{r}
top_songs_features <- get_track_audio_features(top_songs_2022$track.id)
```



```{r}
top_songs_features <- top_songs_features %>% select(id,danceability,energy,
                                                    key, loudness, mode,
                                                    speechiness, acousticness,
                                                    instrumentalness, liveness,
                                                    valence, tempo)
```
```{r}
top_songs_features
```

attach it with the song list
```{r}
top_songs <- left_join(top_songs_2022,top_songs_features,by = c('track.id' = 'id'))
```

```{r}
top_songs
```

Collecting the artists from this playlist

```{r}
artists_top <- top_songs$track.artists
```

```{r}
length(artists_top)
```
```{r}
artists_top[[2]]
```


```{r}
artists <- data.frame(id = c(),name = c())
for (i in 1:length(artists_top)){
  artists <- rbind(artists,(artists_top[[i]] %>% select(id,name)))
}
artists
```
```{r}
artists_freq <- data.frame(table(artists$name, artists$id))
arrange(artists_freq,desc(artists_freq$Freq))
```
```{r}
artists_needed <- artists_freq %>% filter(Freq > 1)
colnames(artists_needed) <- c('name','id','Freq')
artists_needed
```

Getting artists that are similar

```{r}
final_artists <- artists_needed %>% select(id, name)
for (id in artists_needed$id){
  print(id)
  final_artists <- rbind(final_artists,
                         get_related_artists(id) %>% select(id,name))
}
final_artists
```

```{r}
final_artists <- unique(final_artists)
```
```{r}
final_artists
```


Getting the entire library

```{r}
datalists <- vector("list", length = length(final_artists))
artists <- final_artists$id
i=218
for (artist in artists[218:220]){
  dat <- get_artist_audio_features(artist = artist)
  datalists[[i]] <- dat
  i = i+1
}
artist_df <- do.call(rbind, datalists)
```




```{r}
dim(artist_df)
```

This is too large of a dataset. 

```{r}
artists_needed
```
Just picking favourite artists and their related artists

```{r}
fav_artists <- c("Geeta Dutt","Shubha Mudgal","Mohammed Rafi",
                 "Ali Zafar","Begum Akhtar","Mehdi Hassan","Kaavish",
                 "Hemant Kumar","Noor Jehan")
```

```{r}
fav_artists <- final_artists %>% filter(name %in% fav_artists)
```

```{r}
related_artists <- fav_artists %>% select(id, name)
for (id in related_artists$id){
  print(id)
  related_artists <- rbind(related_artists,
                         get_related_artists(id) %>% select(id,name))
}
related_artists
```
```{r}
related_artists <- unique(related_artists)
```

```{r}
related_artists
```
```{r}
dim(df_songs %>% filter(id %in% related_artists$id))
```
```{r}
dim(df_songs)
```
```{r}
dim(df_songs %>% filter(artist_id %in% related_artists$id))
```
```{r}
df <- df_songs %>% filter(artist_id %in% related_artists$id)

```

```{r}
unique(df$artist_name)
```


Top songs and the library to csv files

```{r}
path <- "C:/Users/gadge/OneDrive/Desktop/InfoStats_R/"
filename <- "Top_songs_2022.csv"
write_csv(top_songs,paste(path,filename,sep=""))
```

```{r}
head(df)
```

```{r}
df <- df %>% select(-artists,-album_images,-available_markets )
```

```{r}
filename <- "library_recommender.csv"
write_csv(df,paste(path,filename,sep=""))
```

