---
title: "Final_Project_R"
author: "Uday Gadge"
date: "2022-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Spotify data collection
 
The data is collected through the spotify API. The API is a free resource that spotify provides. Let's use this to collect some data.

The first step is to generate a spotify access token from the client ID and password generated on the spotify developer website.

```{r}
library(spotifyr)
Sys.setenv(SPOTIFY_CLIENT_ID = '39fab145f77749eb9e00395cb262d536')
Sys.setenv(SPOTIFY_CLIENT_SECRET = '6e8c8e372ede44eb9f233a9485f50538')
Sys.setenv(redirect_uri = 'http://localhost:8888/callback')

access_token <- get_spotify_access_token()
```

- Get top 30 artists for every genre
- Get their music libraries
- append for every genre and store it to a csv file

```{r,include = FALSE}
library(dplyr)
library(tidyverse)
path <- "C:/Users/gadge/OneDrive/Desktop/InfoStats_R/"
genres <- c("pop","edm","rap","lounge","classical","alternative","house","Broadway","country","jazz","disco","folk")

for (genre in genres){
  df <- get_genre_artists(genre = genre, limit = 30) %>%
    select(id,name,popularity,genre,followers.total)
  artists <- df$id
  datalists = vector("list", length = length(artists))
  i=1
  for (artist in artists){
    dat <- get_artist_audio_features(artist = artist)
    datalists[[i]] <- dat
    i = i+1
  }
  artist_df <- do.call(rbind, datalists)
  df_artist <- left_join(df,artist_df,by = c("id"="artist_id"))
  df_artist <- df_artist %>% select(-artists,-album_images,-available_markets )
  filename <- paste(genre,".csv")
  write.csv(df_artist,paste(path,filename,sep = ""))
  
}
```
Adding all genres to one data frame


```{r, include = FALSE}
path <- "C:/Users/gadge/OneDrive/Desktop/InfoStats_R/"
genres <- c("pop","edm","rap","lounge","jazz","country","indie","alternative","folk")
i = 1
datalists = vector("list", length = length(genres))
for (genre in genres){
  filename <- paste(genre,".csv")
  genre_df <- read_csv(paste(path, filename, sep = ""))
  datalists[[i]] = genre_df
  i = i+1
}
df <- do.call(rbind, datalists)
```
```{r}
filename <- "Spotify_data.csv"
write_csv(df,paste(path,filename,sep=""))
```

